Complexity analysis and a fully polynomial-time approximation scheme for the college application problem

### 2022-07-01 paper/paper.tex
paragraph about descriptive research; fix bugs in bibliography

@@

To the best of our knowledge, the first systematic study of the college application p·roblem was conducted by this article's first author in his master's thesis (Kapur 2022). A special case of \eqref{headlineproblem} arose in an equilibrium analysis of the United States college admissions market by Fu (2014). However, her econometric task involved clustering the schools such that $m=8$, a scale at which enumeration is tractable. We are interested in solution techniques that scale efficiently in the number of colleges. The present article extends Kapur's key results with a more precise analysis of the problem's computational complexity and a new solution algorithm.

@@

\parskip 0em
\leftskip 2em
\parindent -2em
%Acharya, Mohan S., Asfia Armaan, and Aneeta S. Antony. 2019. ``A Comparison of Regression Models for Prediction of Graduate Admissions.'' In \emph{Second International Conference on Computational Intelligence in Data Science.} \url{https://doi.org/10.1109/ICCIDS.2019.8862140}.
%
%Ashlagi, Itai, and Afshin Nikzad. 2020. ``What Matters in School Choice Tie-Breaking? How Competition Guides Design.'' \emph{Journal of Economic Theory} 190: article no. 105120. \url{https://doi.org/10.1016/j.jet.2020.105120}.
%
%Assad, Arjang. 1985. ``Nested Optimal Policies for Set Functions with Applications to Scheduling.'' \emph{Mathematics of Operations Research} 10 (1): 82--99.
%
%Azevedo, Eduardo and Jacob Leshno. 2016. ``A Supply and Demand Framework for Two-Sided Matching Markets.'' \emph{Journal of Political Economy} 124 (5): 1235--68. \url{https://doi.org/10.1086/687476}. 
%
Badanidiyuru, Ashwinkumar and Jan Vondrák. 2014. ``Fast Algorithms for Maximizing Submodular Functions.'' In \emph{Proceedings of the 2014 Annual ACM--SIAM Symposium on Discrete Algorithms}, 1497--1514. \url{https://doi.org/10.1137/1.9781611973402.110}.

@@

%Cormen, Thomas, Charles Leiserson, and Ronald Rivest. 1990. \emph{Introduction to Algorithms.} Cambridge, MA: The MIT Press.
%
%Dantzig, George B. 1957. ``Discrete-Variable Extremum Problems.'' \emph{Operations Research} 5 (2): 266--88.
%
%Dean, Brian, Michel Goemans, and Jan Vondr\'ak. 2008. ``Approximating the Stochastic Knapsack Problem: The Benefit of Adaptivity.'' \emph{Mathematics of Operations Research} 33 (4): 945--64. \url{https://doi.org/10.1287/moor.1080.0330}.
%
Edmonds, Jack. 1971. ``Matroids and the Greedy Algorithm.'' \emph{Mathematical Programming} 1: 127--36. \url{https://doi.org/10.1007/BF01584082}.  
%
%
%%Fredman, Michael Lawrence and Robert Tarjan. 1987. ``Fibonacci Heaps and Their Uses in Improved Network Optimization Algorithms.'' \emph{Journal of the Association for Computing Machinery} 34 (3): 596--615.
%
Fu, Chao. 2014. ``Equilibrium Tuition, Applications, Admissions, and Enrollment in the College Market.'' \emph{Journal of Political Economy} 122 (2): 225--81. \url{https://doi.org/10.1086/675503}.

@@

%Lim, Daniel Kibum. 2013. ``A Simulation Approach to Predicting College Admissions.'' Master's thesis, University of California Los Angeles. \url{https://escholarship.org/uc/item/8r2695n5}.
%
%Markowitz, Harry. 1952. ``Portfolio Selection.'' \emph{The Journal of Finance} 7 (1): 77--91. \url{https://www.jstor.org/stable/2975974}.
%
%Martello, Silvano and Paolo Toth. 1990. \emph{Knapsack Problems: Algorithms and Computer Implementations.} New York: John Wiley \& Sons. 
%
%Meucci, Attilio. 2005. \emph{Risk and Asset Allocation.} Berlin: Springer-Verlag, 2005. 
%
%% 1 - 1/e inapproximability of submodular maximization
Nemhauser, George and Laurence Wolsey. 1978. ``Best Algorithms for Approximating the Maximum of a Submodular Set Function.'' \emph{Mathematics of Operations Research} 3 (3): 177--88. \url{https://doi.org/10.1287/moor.3.3.177}.

@@

%%Othman, Abraham, Eric Budish, and Tuomas Sandholm. 2010. ``Finding Approximate Competitive Equilibria: Efficient and Fair Course Allocation.'' In \emph{Proceedings of 9th International Conference on Autonomous Agents and Multiagent Systems.} New York: ACM. \url{https://dl.acm.org/doi/abs/10.5555/1838206.1838323}.
%
%Parker, R. Gary and Ronald L. Rardin. 1988. \emph{Discrete Optimization.} San Diego: Academic Press.
%
%Roth, Alvin E. 1982. ``The Economics of Matching: Stability and Incentives.'' \emph{Mathematics of Operations Research} 7 (4): 617--28. \url{https://www.jstor.org/stable/3689483}.
%
%Rozanov, Mark and Arie Tamir. 2020. ``The Nestedness Property of the Convex Ordered Median Location Problem on a Tree.'' \emph{Discrete Optimization} 36: 100581. \url{https://doi.org/10.1016/j.disopt.2020.100581}.
%
%Sklarow, Mark. 2018. \emph{State of the Profession 2018: The 10 Trends Reshaping Independent Educational Consulting.} Technical report, Independent Educational Consultants Association. \url{https://www.iecaonline.com/wp-content/uploads/2020/02/IECA-Current-Trends-2018.pdf}.
%
%Sniedovich, Moshe. 1980. ``Preference Order Stochastic Knapsack Problems: Methodological Issues.'' \emph{The Journal of the Operational Research Society} 31 (11): 1025--32. \url{https://www.jstor.org/stable/2581283}. 
%
%Steinberg, E. and M. S. Parks. 1979. ``A Preference Order Dynamic Program for a Knapsack Problem with Stochastic Rewards.'' \emph{The Journal of the Operational Research Society} 30 (2): 141--47. \url{https://www.jstor.org/stable/3009295}. 
%
%% Another ex of nestedness.
%%Tibshirani, Robert. 1996. ``Regression Shrinkage and Selection via the Lasso.'' \emph{Journal of the Royal Statistical Society,. Series B (Methodological)} 58, no. 1: 267--88. \url{https://www.jstor.org/stable/2346178}.
%
%Van Huizen, Thomas and Rob Alessie. 2019. ``Risk Aversion and Job Mobility.’’ \emph{Journal of Economic Behavior \& Organization} 164: 91--106. \url{https://doi.org/10.1016/j.jebo.2019.01.021}.
%
%Vazirani, Vijay. 2001. \emph{Approximation Algorithms.} Berlin: Springer. 
%
Woeginger, Gerhard. 2005. ``A Comment on Scheduling Two Parallel Machines with Capacity Constraints.'' \emph{Discrete Optimization} 2 (3): 269--72. \url{https://doi.org/10.1016/j.disopt.2005.06.005}.

>>>>

% Move this paragraph to the Alma branch
Most quantitative research on college admissions to date has been descriptive in nature. A standard benchmark for logistic regression techniques, for example, involves estimating students' admissions probabilities from information about their past academic performance (Acharya et al. 2019; Lim 2013). When professional admissions consultants advise their clients on where to apply to college, however, they but typically rely on qualitative heuristics, such as balancing applications between safety and reach schools, rather than optimization models (Peck 2021). (Admissions consultants refer to schools with low utility and high admissions probability as \emph{safety schools}, and those with high utility and low admissions probability as \emph{reach schools.})

One way of thinking about the college application problem is as a static variant of the Pandora's Box problem proposed by Weitzman (1979). In a Pandora's Box formulation of college application, the student applies to schools one by one, each time paying the application fee and observing her admissions outcome after a certain time delay. The problem is to determine an optimal \emph{stopping rule} for when Pandora should halt her college search and accept the best admissions offer she has on hand. Weitzman showed that the optimal policy is to stop searching when the value of the current best offer exceeds the maximum \emph{reservation price,} a statistic that represents the expected value of applying to a new college.

Arguably, the static model considered in this study is more hostile to students than the Pandora's Box problem. If, for example, an unlucky Pandora is rejected from a safety school at an early round of application, then she can compensate for the unexpected loss by pivoting to a more risk-averse application strategy. By contrast, the decisionmaker in our college application problem must commit at the outset to applying to every school in her application portfolio. The admissions process used in the United States can be viewed as the concatenation of both problems: In the fall, students solve \eqref{headlineproblem} and send out a batch of applications. Then, upon observing their admissions outcomes in March, they use the Pandora strategy to pursue additional offers by applying to schools that offer rolling admissions.

@@

To the best of our knowledge, the first systematic study of the college application problem was conducted by this article's first author in his master's thesis (Kapur 2022). A special case of \eqref{headlineproblem} arose in an equilibrium analysis of the United States college admissions market by Fu (2014). However, her econometric task involved clustering the schools such that $m=8$, a problem size that can be solved by enumeration. We are interested in solution techniques that scale efficiently in the number of colleges. The present article extends Kapur's key results with a more precise analysis of the problem's computational complexity and a new solution algorithm.

@@

\parskip 0em
\leftskip 2em
\parindent -2em
Acharya, Mohan S., Asfia Armaan, and Aneeta S. Antony. 2019. ``A Comparison of Regression Models for Prediction of Graduate Admissions.'' In \emph{Second International Conference on Computational Intelligence in Data Science.} \url{https://doi.org/10.1109/ICCIDS.2019.8862140}.

%Ashlagi, Itai, and Afshin Nikzad. 2020. ``What Matters in School Choice Tie-Breaking? How Competition Guides Design.'' \emph{Journal of Economic Theory} 190: article no. 105120. \url{https://doi.org/10.1016/j.jet.2020.105120}.
%
%Assad, Arjang. 1985. ``Nested Optimal Policies for Set Functions with Applications to Scheduling.'' \emph{Mathematics of Operations Research} 10 (1): 82--99.
%
%Azevedo, Eduardo and Jacob Leshno. 2016. ``A Supply and Demand Framework for Two-Sided Matching Markets.'' \emph{Journal of Political Economy} 124 (5): 1235--68. \url{https://doi.org/10.1086/687476}. 
%
Badanidiyuru, Ashwinkumar and Jan Vondrák. 2014. ``Fast Algorithms for Maximizing Submodular Functions.'' In \emph{Proceedings of the 2014 Annual ACM--SIAM Symposium on Discrete Algorithms}, 1497--1514. \url{https://doi.org/10.1137/1.9781611973402.110}.

@@

%Cormen, Thomas, Charles Leiserson, and Ronald Rivest. 1990. \emph{Introduction to Algorithms.} Cambridge, MA: The MIT Press.
%
%Dantzig, George B. 1957. ``Discrete-Variable Extremum Problems.'' \emph{Operations Research} 5 (2): 266--88.
%
%Dean, Brian, Michel Goemans, and Jan Vondr\'ak. 2008. ``Approximating the Stochastic Knapsack Problem: The Benefit of Adaptivity.'' \emph{Mathematics of Operations Research} 33 (4): 945--64. \url{https://doi.org/10.1287/moor.1080.0330}.
%
Edmonds, Jack. 1971. ``Matroids and the Greedy Algorithm.'' \emph{Mathematical Programming} 1: 127--36. \url{https://doi.org/10.1007/BF01584082}.

%Fredman, Michael Lawrence and Robert Tarjan. 1987. ``Fibonacci Heaps and Their Uses in Improved Network Optimization Algorithms.'' \emph{Journal of the Association for Computing Machinery} 34 (3): 596--615.
%
Fu, Chao. 2014. ``Equilibrium Tuition, Applications, Admissions, and Enrollment in the College Market.'' \emph{Journal of Political Economy} 122 (2): 225--81. \url{https://doi.org/10.1086/675503}.

@@

Lim, Daniel Kibum. 2013. ``A Simulation Approach to Predicting College Admissions.'' Master's thesis, University of California Los Angeles. \url{https://escholarship.org/uc/item/8r2695n5}.

%Markowitz, Harry. 1952. ``Portfolio Selection.'' \emph{The Journal of Finance} 7 (1): 77--91. \url{https://www.jstor.org/stable/2975974}.
%
%Martello, Silvano and Paolo Toth. 1990. \emph{Knapsack Problems: Algorithms and Computer Implementations.} New York: John Wiley \& Sons. 
%
%Meucci, Attilio. 2005. \emph{Risk and Asset Allocation.} Berlin: Springer-Verlag, 2005. 
%
%% 1 - 1/e inapproximability of submodular maximization
Nemhauser, George and Laurence Wolsey. 1978. ``Best Algorithms for Approximating the Maximum of a Submodular Set Function.'' \emph{Mathematics of Operations Research} 3 (3): 177--88. \url{https://doi.org/10.1287/moor.3.3.177}.

@@

%%Othman, Abraham, Eric Budish, and Tuomas Sandholm. 2010. ``Finding Approximate Competitive Equilibria: Efficient and Fair Course Allocation.'' In \emph{Proceedings of 9th International Conference on Autonomous Agents and Multiagent Systems.} New York: ACM. \url{https://dl.acm.org/doi/abs/10.5555/1838206.1838323}.
%
%Parker, R. Gary and Ronald L. Rardin. 1988. \emph{Discrete Optimization.} San Diego: Academic Press.

Peck, Timothy. 2021. ``Safety, Target, \& Reach Schools: How to Find the Right Ones.'' CollegeVine Blog, May 7. \url{https://blog.collegevine.com/the-college-list-decoded-safeties-targets-and-reaches/}.

%Roth, Alvin E. 1982. ``The Economics of Matching: Stability and Incentives.'' \emph{Mathematics of Operations Research} 7 (4): 617--28. \url{https://www.jstor.org/stable/3689483}.
%
%Rozanov, Mark and Arie Tamir. 2020. ``The Nestedness Property of the Convex Ordered Median Location Problem on a Tree.'' \emph{Discrete Optimization} 36: 100581. \url{https://doi.org/10.1016/j.disopt.2020.100581}.
%
%Sklarow, Mark. 2018. \emph{State of the Profession 2018: The 10 Trends Reshaping Independent Educational Consulting.} Technical report, Independent Educational Consultants Association. \url{https://www.iecaonline.com/wp-content/uploads/2020/02/IECA-Current-Trends-2018.pdf}.
%
%Sniedovich, Moshe. 1980. ``Preference Order Stochastic Knapsack Problems: Methodological Issues.'' \emph{The Journal of the Operational Research Society} 31 (11): 1025--32. \url{https://www.jstor.org/stable/2581283}. 
%
%Steinberg, E. and M. S. Parks. 1979. ``A Preference Order Dynamic Program for a Knapsack Problem with Stochastic Rewards.'' \emph{The Journal of the Operational Research Society} 30 (2): 141--47. \url{https://www.jstor.org/stable/3009295}. 
%
%% Another ex of nestedness.
%%Tibshirani, Robert. 1996. ``Regression Shrinkage and Selection via the Lasso.'' \emph{Journal of the Royal Statistical Society,. Series B (Methodological)} 58, no. 1: 267--88. \url{https://www.jstor.org/stable/2346178}.
%
%Van Huizen, Thomas and Rob Alessie. 2019. ``Risk Aversion and Job Mobility.’’ \emph{Journal of Economic Behavior \& Organization} 164: 91--106. \url{https://doi.org/10.1016/j.jebo.2019.01.021}.
%
%Vazirani, Vijay. 2001. \emph{Approximation Algorithms.} Berlin: Springer. 
%
Woeginger, Gerhard. 2005. ``A Comment on Scheduling Two Parallel Machines with Capacity Constraints.'' \emph{Discrete Optimization} 2 (3): 269--72. \url{https://doi.org/10.1016/j.disopt.2005.06.005}.
